Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-21 tex/draft_noComments.tex
Initial commit

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing (*weak word) accuracy (*say 2nd clause better). The SM consists of 12 spin-1/2 particles, the fermions, that make up all observed matter, 12 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces (*maybe put in pre-EWSB terms instead?), and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group SU(3)cxSU(2)LxU(1)Y, where SU(3)c is associated with color charge, SU(2)L is associated with weak isospin, and U(1)Y is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with SU(3)c, 3 W bosons associated with SU(2)L, and 1 B boson associated with U(1)Y. The fermions are catagorized by which representation of each group they fit into. For example, SU(3)c distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and SU(2)L distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, different values of weak hypercharge further distignuish SM particles according to a linear combination of their weak isospin and electric charge (*is there some more fundamental way to describe Y?).

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks SU(2)LxU(1)Y to U(1)EM while leaving SU(3)c unscathed. Once the dust settles, the B and 3 W bosons have been mixed to produce the W+, W-, Z bosons that mediate the weak force and the photon that mediates electromagnetism. 3 of the 4 generators are broken in the process, leaving 3 extra degrees of freedom that ultimately become the masses of the W+, W-, and Z bosons while the remaining symmetry is associated with electric charge. Furthermore, expanding the Higgs field about its VEV produces mass terms for the quarks and charged leptons, thus giving mass to all massive SM particles. (*probably show a little math)

    (*kinda funny wording throughout. Maybe it would be better to trade some words for math?)
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections. As the only fundamental scalar in SM, the Higgs is uniquely affected by these corrections. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms. Here are a few (*cite these): (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad (*probably a better way to say 4). Although these statements are all trying to get at the same fundamental notion, I find (3) and (4) to be most compelling.

    (*cite guidice naturally and t'hooft naturalness)
    Let's look at (3). In the SM, the small parameter of interest is the Higgs mass, which has no custodial symmetry to hold it down. This situation contrasts that of the fermions and gauge bosons, whose masses are kept small by custodial symmetries. That is, the SM would be more symmetric (*weird wording?) if fermions and gauge bosons were massless. Specifically, massless fermions would have an extra U(1)xU(1) chiral symmetry, and massless vector bosons would have an extra gauge symmetry (*?!?!?!). The Higgs, a fundamental scalar, has no such symmetry (*repetative) so we deem its low mass to be unnatural. (*explain why fundamental scalars have no custodial symmetry? I certainly better learn why.)

@@

    (*probably doesn't need to be in final paper, but I'd like to check my basic understanding with Jamie, et al)
    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally ~60 to ~120 orders of magnitude smaller than the predicted value (*predicted how?). While these issues deserve serious attention, they will not be considered further here.

    (*talk about anthropic principle or other arguments against?)
    (*I got tired of reading this section and haven't edited this paragraph at all yet)
    There are a few proposed solutions to the SM naturalness problem. Supersymmetry (SUSY) is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence. (*Hidden valley, dark sector, little Higgs?)

\section{SUSY and LLPs}
    (*is there a way to move the naming stuff elsewhere or shorten it?)
    SUSY introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between fermions and bosons. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson (*and an extra Higgs field - can I gloss over that?). The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles. Because fermion loops come with a relative minus sign (*there must be a better way to say that), every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. If supersymmetry were exact, the bare Higgs mass would be uncorrected to all orders in perturbation theory.

@@

    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and (*and/or?) produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict LLPs. That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point, which includes everything from particles that decay at (*give ballparky number (~100um?)) to particles that propagate through the entire detector.

    In both the SM and in BSM theories, LLPs result from one of the following: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, the electron and proton are the lightest particles carrying electric charge and baryon number, respectively, which leads to their absolute stability through charge and baryon number conservation. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay (*also mention that they're both weak decays?). (*Mention neutrino, pion, kaon, whatever other bound state I might be forgetting?)

    In SUSY models, LLPs arise for the same basic reasons, and many different SUSY models meet different LLP criteria in different ways. The most common (*based on my assumptions) SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from predicting proton decay. In models with R-parity, the SM particles have R-parity +1 (*double check this explanation) and their superpartners have R-parity -1. Conserving overall R-parity (*specifically the product; does overall mean that?) at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is an absolutely stable LLP. The second phenomenon is the main motivation behind using missing energy as the typical SUSY seach signature.

    In models with R-parity violation (RPV), LLPs can be produced when the RPV terms have small coupling constants, which leads to long-lived, but not perfectly stable, LSPs (*any way non-LSPs could be long-lived for this reason too?). Luckily, RPV models can still avoid proton decay by only allowing terms that violate either baryon number or lepton number but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs(*do I need to explain GMSB here?). (*cite dimopoulos low energy)

    LLPs from highly-virtual intermediate particles can occur in models such as split (or mini-split) SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. (*cite kilian) (*mention that split throws out naturalness motivation?)

    SUSY models with degenerate sparticle spectra can have increased sparticle lifetimes from the reduced phase space for decays. In some AMSB models (*be more specific), for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino's lifetime, then, is increased by the lack of phase space in which to decay. (*cite R\&S out of this world)
    (*merge last 2 paragraphs under 'mass spectra lead to LLP' concept?)
    (*never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?)

\section{The CMS Detector}
\subsection{Overview}
        (*cite cms experiment at the cern lhc)
          The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from 14 TeV proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a 4-T superconducting solenoid. All together, CMS is 21.6 m long, 14.6 m across, and weighs approximately 12500 tons. The subdetectors that make up CMS are described below.

\subsection{Tracker}
        (*cite CMS TDR, jinst 12 c02033)
        (*maybe I should trade some number listing for talking about how silicon detectors work)
        (*what about readout, cooling, etc?)
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint (*too flowery?) primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux (*specify) requires smaller pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size (*look up) depending on particle flux, the SST covers an active area of roughly 196 m\^2. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics (*right language?).

\subsection{Electromagnetic Calorimeter}
        (*cite CMS TDR)
        (*talk about readout?)
        (*More in the way of concept/how stuff works?)
        Moving outward, the electromagnetic calorimeter (ECAL) follows the inner tracker. As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad(*learn context for this number)) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields (*say B-field part better).

        The barrel section (EB) extends radially from 129 to 177 (*look up) cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that particle trajectories would be likely to align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

        The crystals in each endcap section (EE) are arranged in an x-y grid that starts at |z| = 315 cm and covers a pseudorapidity range of 1.479 to 3.0. Additionally, a preshower detector, which consists of two planes of silicon strip detectors and a lead absorber, sits just inside the grid of lead tungstate crystals. (*Say what the preshower is for? Is it for pi0 rejection? If so, how and why?)

\subsection{Hadronic Calorimeter}
        (*cite CMS TDR)
        (*talk about readout?)
        The hadronic calorimeter (HCAL) uses 3.7mm plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers so it can eventually be read out by hybrid photodiodes.

        The barrel section (HB) spans a pseudorapidty range of -1.4 to 1.4 and has 2304 towers (*check), each of which contains 14 (*check) brass plates and covers an area of 0.087 x 0.087 in eta-phi (*get a feel for this area). In addition, an extra layer (or two at eta = 0) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidty range of -1.26 to 1.26 and increases the minimum effective HCAL interaction length to greater than 11.8.

        Each endcap consists spans a pseudorapidity range of 1.3 to 3.0 with 14 towers in eta and 5 to 10 degree phi segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits 11.2 m from the interaction pipe in a pseudorapidty range of 3.0 to 5.0. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. (*say why HF is useful or mention crazy flux?)

\subsection{Solenoid}
        (*cite cms experiment at the cern lhc)
        CMS employs a 4 T superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of 2168 turns of Nb-Ti superconducting cable and has an inner diameter of 6 m and a length of 12.5 m. The flux returns through an iron yoke that also houses the muon system. The significant (*quantify?) bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of 10\% for muons up to 1 TeV.

\subsection{Muon System}
        (*cite CMS TDR)
        (*Specifically mention CMS design specs to meet physics goals?)
        The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in both the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines (*quantify or show std plot?).

        In the barrel detector (BD), four layers of DTs and and RPCs are embedded in the solenoid return yoke at 4.0, 4.9, 5.9, and 7.9 m. (*Say how DT works?) The maximum drift length is 2 cm and the single point resolution is 200 um. When reconstructing a vector describing the muon momentum, the phi precision is better than 100 um in position and 1 mrad in direction.

        Each muon endcap detector (ME) is composed of 234 CSCs and 4 RPC stations (*figure out exactly what RPC station means and say that instead). Each CSC has six gas gaps with radial cathode strips and nearly perpendicualar planes of anode wires. The wire signal from the ionization-induced electron avalanch is fast enough to use in the L1 trigger, while the slower (*quantify) cathode strip signal provides precise (*quantify) position measurements. All in all, the CSC spacial resolution is approximately 200 um and the angular resolution is approximately 10 mrad.

\subsection{Trigger}
        (*cite CMS TDR, jinst 12 c03021)
        (*Is entire detector read out at L1?)
        (*Mention fpgas in L1 or more details about HLT?)
        (*Something about particle flow?)
        (*Sloppy, but I'm getting tired)
        The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). The HLT uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, ET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to reduce the rate to the desired < 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
        LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to veer off the traditional analysis path in terms of object reconstruction (*really reconstruction? Maybe object definition?), background estimation, and which data tier to use (*can I say this more generally?). This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

        Displaced vertices and disappearing tracks both look for particles that are unstable on detector timescales (*quantify?). Displaced vertices searches are sensitive to LLPs that decay to visible (*detectable? some other word?) particles in the inner volume (*specify) of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point (*wordy). Disappearing tracks searches are sensitive to LLPs that decay to invisible (*neutral? some other word?) particles along with particles that are too soft to be reconstructed. Together, these two types of analyses cover something something parameter space of LLPs that decay on detector timescales.

        Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles (*redundant), which normally (*quantify) have beta > 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system (*too specific for overview?) that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
            (*cite CONF-2017-026)
            LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified by their displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET (*has MET been defined yet?) and large track multiplicity. The search is sensitive to long-lived gluinos with lifetimes in the 20 ps to 10 ns range.

            This search uses a simplified model in the split SUSY paradigm as a benchmark. In the simplified model, gluinos are kinematically accessible while squarks are not. The relevant process, then, is pair produced gluinos that decay to a quark and virtual squark, which then decays to a quark and the LSP, which is presumed to be a neutralino (*diagram?). Because the squark mass is so high, the gluino lives long enough to form an R-hadron (*first use. Define here if not in LLP section) with SM quarks and propogate some measureable distance away from the IP before decaying.

            Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0, and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs. (*should I talk about reconstruction efficiencies?)

@@

            After estimating 0.02 +- 0.02 background events, exactly 0 are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime. For an LSP neutralino mass of 100 GeV, the gluino mass limit is approximately 2000 GeV for lifetimes in the 20 ps to 10 ns range. (*Include plots)

>>>>

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and forces (except gravity) and is experimentally verified to astonishing accuracy. The SM consists of 12 spin-1/2 particles, the fermions, that make up all observed matter, 5 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces, and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    The SM particles fit into multiplets of three symmetry groups, which makes the SM an SU(3)cxSU(2)LxSU(1)Y theory. These three symmetry groups correspond to color, the charge associated with the strong nuclear force; weak isospin, the charge associated with the left-handed chiral doublet something something; and weak hypercharge, the something something something, respectively.

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks the SU(2)LxU(1)Y symmetry to a U(1)EM symmetry. Breaking 3 of the 4 generators leaves 3 extra degrees of freedom, which ultimately become the masses of the W+, W-, and Z bosons, while the remaining degree of freedom is associated with electric charge and the massless photon. Expanding the Higgs field about its VEV and picking a gauge produces quark and charged lepton mass terms in the SM Lagrangian. In this way, spontaneous symmetry breaking of the Higgs field gives mass to all massive SM particles.

    Like the mass of all SM particles, the Higgs mass is affected by interactions with virtual particles through loop processes. As the only fundamental scalar in the theory, however, the Higgs sees corrections to its mass that depend quadratically on the cutoff scale of the SM. To show this dependence on loop corrections, SM masses are often written as power series, where the first term is the bare (uncorrected) mass and the following terms are due to loop corrections. The next section will explore the concept of naturalness, a major motivating principle in beyond the SM (BSM) physics, that is directly tied to the relationship between a particle's bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms. Here are a few: (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad. Although these statements are all trying to get at the same fundamental notion, I find (3) and (4) to be most compelling.

    Let's look at (3). In the SM, the small parameter of interest is the Higgs mass, which has no custodial symmetry to hold it down. This situation contrasts that of the fermions and gauge bosons, whose masses are kept small by custodial symmetries. That is, the SM would be more symmetric if fermions and gauge bosons were massless. Specifically, massless fermions would have an extra U(1)xU(1) chiral symmetry, and massless vector bosons would have an extra gauge symmetry. The Higgs, a fundamental scalar, has no such symmetry so we deem its low mass to be unnatural.

@@

    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally ~60 to ~120 orders of magnitude smaller than the predicted value. While these issues deserve serious attention, they will not be considered further here.

    There are a few proposed solutions to the SM naturalness problem. Supersymmetry (SUSY) is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence.

\section{SUSY and LLPs}
    SUSY introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between fermions and bosons. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson. The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles. Because fermion loops come with a relative minus sign, every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. If supersymmetry were exact, the bare Higgs mass would be uncorrected to all orders in perturbation theory.

@@

    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict LLPs. That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point, which includes everything from particles that decay at to particles that propagate through the entire detector.

    In both the SM and in BSM theories, LLPs result from one of the following: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, the electron and proton are the lightest particles carrying electric charge and baryon number, respectively, which leads to their absolute stability through charge and baryon number conservation. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay.

    In SUSY models, LLPs arise for the same basic reasons, and many different SUSY models meet different LLP criteria in different ways. The most common SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from predicting proton decay. In models with R-parity, the SM particles have R-parity +1 and their superpartners have R-parity -1. Conserving overall R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is an absolutely stable LLP. The second phenomenon is the main motivation behind using missing energy as the typical SUSY seach signature.

    In models with R-parity violation (RPV), LLPs can be produced when the RPV terms have small coupling constants, which leads to long-lived, but not perfectly stable, LSPs. Luckily, RPV models can still avoid proton decay by only allowing terms that violate either baryon number or lepton number but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs.

    LLPs from highly-virtual intermediate particles can occur in models such as split (or mini-split) SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark.

    SUSY models with degenerate sparticle spectra can have increased sparticle lifetimes from the reduced phase space for decays. In some AMSB models, for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino's lifetime, then, is increased by the lack of phase space in which to decay.

\section{The CMS Detector}
\subsection{Overview}
          The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from 14 TeV proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a 4-T superconducting solenoid. All together, CMS is 21.6 m long, 14.6 m across, and weighs approximately 12500 tons. The subdetectors that make up CMS are described below.

\subsection{Tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux requires smaller pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size depending on particle flux, the SST covers an active area of roughly 196 m\^2. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics.

\subsection{Electromagnetic Calorimeter}
        Moving outward, the electromagnetic calorimeter (ECAL) follows the inner tracker. As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields.

        The barrel section (EB) extends radially from 129 to 177 cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that particle trajectories would be likely to align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

        The crystals in each endcap section (EE) are arranged in an x-y grid that starts at |z| = 315 cm and covers a pseudorapidity range of 1.479 to 3.0. Additionally, a preshower detector, which consists of two planes of silicon strip detectors and a lead absorber, sits just inside the grid of lead tungstate crystals.

\subsection{Hadronic Calorimeter}
        The hadronic calorimeter (HCAL) uses 3.7mm plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers so it can eventually be read out by hybrid photodiodes.

        The barrel section (HB) spans a pseudorapidty range of -1.4 to 1.4 and has 2304 towers, each of which contains 14 brass plates and covers an area of 0.087 x 0.087 in eta-phi. In addition, an extra layer (or two at eta = 0) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidty range of -1.26 to 1.26 and increases the minimum effective HCAL interaction length to greater than 11.8.

        Each endcap consists spans a pseudorapidity range of 1.3 to 3.0 with 14 towers in eta and 5 to 10 degree phi segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits 11.2 m from the interaction pipe in a pseudorapidty range of 3.0 to 5.0. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline.

\subsection{Solenoid}
        CMS employs a 4 T superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of 2168 turns of Nb-Ti superconducting cable and has an inner diameter of 6 m and a length of 12.5 m. The flux returns through an iron yoke that also houses the muon system. The significant bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of 10\% for muons up to 1 TeV.

\subsection{Muon System}
        The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in both the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines.

        In the barrel detector (BD), four layers of DTs and and RPCs are embedded in the solenoid return yoke at 4.0, 4.9, 5.9, and 7.9 m. The maximum drift length is 2 cm and the single point resolution is 200 um. When reconstructing a vector describing the muon momentum, the phi precision is better than 100 um in position and 1 mrad in direction.

        Each muon endcap detector (ME) is composed of 234 CSCs and 4 RPC stations. Each CSC has six gas gaps with radial cathode strips and nearly perpendicualar planes of anode wires. The wire signal from the ionization-induced electron avalanch is fast enough to use in the L1 trigger, while the slower cathode strip signal provides precise position measurements. All in all, the CSC spacial resolution is approximately 200 um and the angular resolution is approximately 10 mrad.

\subsection{Trigger}
        The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). The HLT uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, ET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to reduce the rate to the desired < 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
        LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to veer off the traditional analysis path in terms of object reconstruction, background estimation, and which data tier to use. This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

        Displaced vertices and disappearing tracks both look for particles that are unstable on detector timescales. Displaced vertices searches are sensitive to LLPs that decay to visible particles in the inner volume of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point. Disappearing tracks searches are sensitive to LLPs that decay to invisible particles along with particles that are too soft to be reconstructed. Together, these two types of analyses cover something something parameter space of LLPs that decay on detector timescales.

        Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles, which normally have beta > 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
            LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified by their displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity. The search is sensitive to long-lived gluinos with lifetimes in the 20 ps to 10 ns range.

            This search uses a simplified model in the split SUSY paradigm as a benchmark. In the simplified model, gluinos are kinematically accessible while squarks are not. The relevant process, then, is pair produced gluinos that decay to a quark and virtual squark, which then decays to a quark and the LSP, which is presumed to be a neutralino. Because the squark mass is so high, the gluino lives long enough to form an R-hadron with SM quarks and propogate some measureable distance away from the IP before decaying.

            Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0, and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs.

@@

            After estimating 0.02 +- 0.02 background events, exactly 0 are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime. For an LSP neutralino mass of 100 GeV, the gluino mass limit is approximately 2000 GeV for lifetimes in the 20 ps to 10 ns range.

\subsection{Disappearing Tracks}
            Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos, such as many AMSB models, will have a chargino NLSP with an observable lifetime (0.2 ns is a typical estimate) that eventaully decays to the neutralino LSP and, commonly, a pion. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be constructed, which results in a disappearing track.

            Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 and CMS before 2017 should increase sensitivity to decays at shorter radii. The ALTAS search presented here is sensitie to LLP lifetimes of approximated 10 ps to 10 ns.
